---
title: "Challenge_4: Intro to Visulization: Univariate and Multivariate Graphs"
author: "Sagnik Chand"
description: "This is our fourth challenge and we will utilise ggplot to generate visualizations in R"
date: "10/18/2023"
format:
  html:
    df-print: paged
    css: "styles.css"
    embed-resources: true
    self-contained-math: true
categories:
  - weekly_challenges
  - challenge_4
---

**Make sure you change the author's name in the above YAML header.**

## Setup

If you have not installed the following packages, please install them before loading them.

```{r}
#| label: setup
#| warning: false
#| message: false

library(tidyverse)
library(readxl)
library(haven) #for loading other datafiles (SAS, STATA, SPSS, etc.)
library(stringr) # if you have not installed this package, please install it.
library(ggplot2) # if you have not installed this package, please install it.
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Challenge Overview

In this challenge, we will practice with the data we worked on in the previous challenges and the data you choose to do some simple data visualizations using the `ggplot2` package.

There will be coding components and writing components. Please read the instructions for each part and complete your challenges.

## Datasets

-   Part 1 the ESS_Polity Data (created in Challenge#3) ⭐⭐
-   Part 2: the Australia Data⭐⭐
-   Part 3: see [Part 3. Practice plotting with a dataset of your choice (25% of the total grade)](#part-3.-practice-plotting-with-a-dataset-of-your-choice-25-of-the-total-grade). For online platforms of free data, see [Appendix: sources for data to be used in Part 3].

Find the `_data` folder, then read the datasets using the correct R command.

## Part 1. Univariate and Multivariate Graphs (45% of the total grade)

We have been working with these two data in the previous three challenges. Suppose we have a research project that studies European citizens' social behaviors and public opinions, and we are interested in how the countries that respondents live in influence their behavior and opinion. In this challenge, let's work with the combined dataset *ESS_Polity* and create some visualizations.

1.  **Read the combined data you created last time. (2.5%)**

```{r}
# Reading in the data set and including a sanity check.

Ess_Polity <- read_csv("Ess_Polity.csv")
head(Ess_Polity)

```

2.  **Suppose we are interested in the central tendencies and distributions of the following variables. At the individual level:** *age, male, edu, income_10,* and *vote*. **At the country level**: *democ*.

    \(1\) Recode the "vote" column: if the value is 1, recode it as 1; if the value is 2, recode it as 0; if the value is 3, recode it as NA. **Make sure to include a sanity check for the recoded data. (2.5%)**

    ```{r}
    # Recoding the dataset as required and including a sanity check.

    Ess_Polity <- Ess_Polity %>% 
      mutate(vote = recode(vote,
                    "1" = "1",
                    "2" = "0",
                    '3' = "NA"))

    unique(Ess_Polity$vote)
    ```

    \(2\) For each of the five variables (*age, edu, income_10,* *vote,* and *democ)*, please choose an appropriate type of univariate graph to plot the central tendencies and distribution of the variables. Explain why you choose this type of graph to present a particular variable (for example: "For example, I use a histogram to plot *age* because it is a continuous numeric variable"). **(25%)**

    **(Note: You should use at least two types of univariate graphs covered in the lecture.)**

    ```{r}
      
    # Age Plot (I used a histogram to plot age because it is a continuous numeric variable. In this example, a bin size between 4 and 8 will show a better pattern.)
      
      ggplot(Ess_Polity, aes(x = age))+
      geom_histogram(binwidth = 4)

    # Education Plot (I used a bar chart to plot edu because it is a categorical variable having high numeric frequencies)
      
      ggplot(Ess_Polity, aes(x = edu))+
      geom_bar()
      
    # Income Plot (I used a boxplot to plot income because it is effective to show central tendency(Median, Distribution and Quartile) and the distribution of data)
      
      ggplot(Ess_Polity, aes(x = income_10))+
      geom_boxplot()+
      labs(x = "Income")

    # Vote Plot (I used a bar chart to plot vote because it is a categorical variable)
      
      ggplot(Ess_Polity, aes(x = vote))+
        geom_bar()                      # The last NA column represents NA values without recoding. 

    # Democ Plot (I used a bar chart to plot democ because it is a categorical variable with 5 categories)
      
      ggplot(Ess_Polity, aes(x = democ))+
        geom_bar()

    ```

3.  **Suppose we want to test two hypotheses on the relationships of two pairs of variables**. **Please use the appropriate type of graphs we learned to visualize these two pairs of variables. Briefly describe the graph you plot, and answer: Does the graph we create from the data support the hypothesis?**

    \(1\) Hypothesis#1: The more years of education (edu) a person completed, the higher income (income_10) they earn. **(7.5%)**

    Ans: I used a scatter plot and added another layer of smooth linear regression line on to it to plot the relationship in between education and income. According to the graph, our hypothesis that "The more years of education (edu) a person completed, the higher income (income_10) they earn" is proven correct.

    ```{r}
    # Education vs Income plot.

    ggplot(Ess_Polity, aes(x = income_10, fill = as.factor(edu)))+
      geom_bar()
      
    ```

    \(2\) Hypothesis#2: There is a gender disparity (male) in voting behavior (vote). (Either men are more likely to vote, or women are more likely to vote). **(7.5%)**

    Ans: I utilized a grouped bar plot to visualize gender disparities in voting behavior, specifically comparing men and women. By employing the "dodge" positioning, the plot facilitates a straightforward comparison between the two genders. The graphical representation supports the validity of Hypothesis #2, revealing a notable gender-based divergence in voting patterns. Notably, the data suggests that women exhibit a higher likelihood of voting compared to men.

    ```{r}
    # Gender vs Vote disparity graph. (Removing all NA values; along with the votes recoded as 'NA')

    Ess_Polity %>% 
      select(male, vote) %>% 
      filter(!is.na(male)) %>%
      filter(!is.na(vote)) %>%
      filter(vote != "NA") %>%
      mutate(male = recode(male,
                           "0" = "Female",
                           "1" = "Male")) %>%
      ggplot(aes(x = vote, fill = as.factor(male)))+
      geom_bar(position = "dodge", stat = "count")+
      labs(x = "Vote", y = "Count", fill = "Gender")

    ```

## Part 2. Comparing between Partial and Whole, and among Groups (30% of the total grade)

In this part, we will use the clean version of the Australian public opinion poll on Same-Sex Marriage to generate graphs and plots. **You may need to do the data transformation or mutation needed to help graphing.**

1.  Read in data. **(2.5%)**

    ```{r}
    # Reading the data set and applying necessary transformations.

    Aus <- read_csv("australian_data.csv")
    head(Aus)

    Aus_long <- pivot_longer(Aus, 2:5, names_to = "Opinion", values_to = "Count")
    head(Aus_long)
    ```

2.  Use a barplot to graph the Australian data based on their responses: yes, no, illegible, and no response. The y-axis should be the count of responses, and each response should be represented by one individual bar (so there should be four bars). **(7.5%)**

    (you can use either geom_bar() or geom_col())

    ```{r}
    # Australian data based on their responses.

    Aus_long %>% 
      group_by(Opinion) %>% 
      summarise( Sum = sum(Count)) %>% 
      ggplot(aes(x = Opinion, y = Sum))+
      geom_bar(stat = "identity")+
      scale_y_continuous(labels = function(x) format(x, scientific = FALSE))

    ```

3.  The previous graph only shows the difference in amount. Let's create a stacked-to-100% barplot to show the proportion of each of the four responses (by % of the total response). **(7.5%)**

    (you can use either geom_bar() or geom_col())

    ```{r}
    # A stacked bar plot showing proportion of each of the four responses.

    Aus_long %>% 
      group_by(Opinion) %>% 
      summarise( Sum = sum(Count)) %>% 
      mutate( Percentage = Sum/sum(Sum)*100) %>% 
      ggplot(aes(x = 1, y = Percentage , fill = Opinion))+
      geom_bar(stat = "identity", position = "stack", width = 0.1)+
      scale_y_continuous(labels = function(x) format(x, scientific = FALSE))

    ```

4.  Let's see if there's a relationship between Division and Response - that is, are certain divisions more likely to respond one way compared to other divisions? Again, we will use barplot(s) to present the visualization. **(12.5%)**

    (you can use either geom_bar() or geom_col())

    ```{r}
    # Relationship between Division and Response.
       
    Aus_long %>% 
      group_by(Division, Opinion) %>% 
      summarise(Count = sum(Count)) %>% 
      mutate(Percentage = Count/sum(Count)*100) %>% 
      ggplot(aes(x = Division, y = Percentage,  fill = Opinion))+
      geom_bar(position = "stack", stat = "identity")+
      coord_flip()

    ```

## Part 3. Practice plotting with a dataset of your choice (25% of the total grade) {#part-3.-practice-plotting-with-a-dataset-of-your-choice-25-of-the-total-grade}

In this part, you will choose data of your interests for graphing and plotting. This data can be tidy/ready-to-be-used or raw data that needs cleaning. If the data is very large (for example, more than 20 columns), you should definitely subset the data by selecting less than 10 variables of your interests to avoid taking too much room in your R memory.

1.  Include a link to the data page (this page should include the introduction or description and the link to download this dataset). **(2%)**

Ans: The link to the Kaggle data page = <https://www.kaggle.com/datasets/mohamedhanyyy/top-football-leagues-scorers?select=Data.csv>

2.  Read the data you choose and briefly answer the following questions. (Optional: you may need to subset, clean, and transform the data if necessary). **(8%)**

    ```{r}
    #type of your code/command here.
    Fut_Score <- read_csv("Football_Scorers.csv")

    head(Fut_Score)
    View(Fut_Score)
    dim(Fut_Score)
    ```

    \(1\) what is the structure (dimension) of the data;

    Ans: This data set contains 660 rows and 15 columns.

    \(2\) what is the unit of observation?

    Ans: The unit of observation is individual football players from different leagues, who were top scorers in the year 2016.

    \(3\) what does each column mean in this data?

    Ans: Each column defines different variables for a particular football player in the season of 2016 such as the league, club and country name, name of the player, number of matches played, goals scored, shots taken, expected goal rate, minutes played, and so on.

3.  Choose two columns/variables of your interests. Plot one univariate graph for each of the variables. **(5%)**

    ```{r}
    # Plotting number of Goals. (I used a boxplot to show the central tendency and distribution of number of goals scored by each player)

    Fut_Score %>% 
      ggplot(aes(Goals))+
      geom_boxplot()

    # Plotting number of shots taken. (I used a histogram to show the continuous numeric frequency of number of shots taken by players)

    Fut_Score %>% 
      ggplot(aes(Shots))+
      geom_histogram(binwidth = 10.36)
    
    ```

<!-- -->

4.  Choose a pair of variables you suspect or hypothesize may be correlated and a graph (scatter plot or barplot) using them. Based on the visual evidence, do you see any potential correlation between the two variables **(10%)**

    Ans: I hypothesize that the number of goals scored by a player would increase with the increase in the expected goal ratio. Based on the Visual evidence, there is a potential correlation between the two variables. When there is an increase in the expected goal percentage, the number of goals scored by that player also increases.

    ```{r}
    # Plotting Goals vs Expected Goals ratio using a scatter plot.

    Fut_Score %>% 
      ggplot(aes(x = Goals, y = xG))+
      geom_point()+
      geom_smooth()+
      labs(title = "Goals vs Expected Goals", y = "Expected Goals ratio")

    ```

## Appendix: sources for data to be used in Part 3

**Here are some online sources and popular Online Dataset Hub:**

1.  Many US governments (usually at the federal and state levels),  bureaus, and departments have open data archives on their websites, allowing the public to access, download, and use them. Just use Google to search for them.

<!-- -->

2.  [**The Harvard Dataverse Repository**](https://dataverse.harvard.edu/) is a free data repository open to all researchers from any discipline, inside and outside the Harvard community, where you can share, archive, cite, access, and explore research data. Each individual Dataverse collection is a customizable collection of datasets (or a virtual repository) for organizing, managing, and showcasing datasets.

<!-- -->

3.  [**Inter-university Consortium for Political and Social Research (ICPSR)**](https://www.icpsr.umich.edu/web/pages/ICPSR/index.html) of the University of Michigan-Ann Arbor provides leadership and training in data access, curation, and methods of analysis for the social science research community. 

<!-- -->

4.  **UN: <https://data.un.org/>**

<!-- -->

5.  [**OECD Data**](https://data.oecd.org/):  economic and development data of the most developed countries in the world.

<!-- -->

6.  The upper five sources are mainly for social science data; **there is another very big community and open data archives for machine-learning and data science: [Kaggle](https://www.kaggle.com/datasets).**
